\Title

### 2023-02-07 Grimaldi-etal-BiolCybernetics.tex
intro : pass

The response of a biological neuron depends on the precise timing of afferent spikes. This temporal aspect of the neural code is essential to understand information processing in neurobiology and applies particularly well to the output of neuromorphic hardware such as event-based cameras. However, most artificial neural models do not take advantage of this important temporal dimension of the neural code. Inspired by this neuroscientific observation, we develop a model for the efficient detection of temporal spiking motifs based on a layer of spiking neurons with heterogeneous synaptic delays, which we apply to the computer vision task of motion detection. Indeed, the diversity of synaptic delays on the dendritic tree allows for the synchronization of synaptic inputs as they reach the basal dendritic tree. We show that this can be formalized as a time-invariant logistic regression that can be trained on labeled data. We apply this model to solve the specific computer vision problem of motion detection. We demonstrate its application to synthetic naturalistic videos transformed into event streams similar to the output of event-based cameras. In particular, we quantify how the accuracy of the model can vary as a function of the overall computational load. This end-to-end, event-driven computational building block could improve the performance of future spiking neural network (SNN) algorithms and neuromorphic chips.

@@

The human brain has the remarkable property of being able to react at any time while consuming a reasonable amount of energy, about tens of watts. This system is the result of millions of years of natural selection, and a striking difference between biological and artificial neural networks is the representation they use. Convolutional neural networks (CNNs), for example, represent the flow of information from one layer to another as tensors, storing visual information densely across the visual topography, with different properties represented in different channels. These networks are known to mimic many properties of biological systems, such that each can be attributed a ``Barin Score''~\citep{schrimpf_brain-score_2020}, yet this score takes no account of speed or energy consumption. CNNs have achieved state-of-the-art performance for some computer vision tasks, such as image recognition. However, they do not take advantage of the dynamics inherent in the way we perceive our natural environment. On the other hand, in the vast majority of biological neural networks, information is represented as \emph{spikes}, i.e. prototypical all-or-nothing events whose only parameters are their timing and the address of the neuron that emitted the spike~\citep{paugam-moisy_computing_2012}. The third generation of artificial neural networks, called Spiking Neural Networks (SNNs), incorporate this temporal dimension into how they compute. By providing a better model of biological neurons, these new algorithms are of interest to computational neuroscience. Some approaches have developed normative models of SNNs that aim to have applications in machine learning. One of these is the SpikeNet algorithm, which uses a purely temporal approach by encoding information using one spike per neuron~\citep{delorme_spikenet_1999}. Another type of SNN that uses accurate timing of spikes attempts to determine the structure of the network to minimize a cost function. This was implemented in the SpikeProp algorithm~\citep{bohte_error-backpropagation_2002} and has been recently extended in gradient-based methods. This surrogate gradient method is now widely used in methods that attempt to transfer the performance of CNNs to SNNs~\citep{zenke_remarkable_2021}. However, the performance of SNNs still lags behind that of firing rate based networks. The question of the advantage of using spikes in machine learning and computer vision remains open.

In a recent review paper, we reported experimental evidence for the presence of precise spiking motifs embedded in recordings from biological neural tissues~\cite{grimaldi_precise_2023}. These spatio-temporal motifs of neural activity may be useful representations used by neurons to perform computations for various cognitive tasks. In 1982, Abeles asked whether the role of cortical neurons is the integration of synaptic inputs or rather the detection of coincidences in temporal spiking motifs~\citep{abeles_role_1982}. While the first possibility favors the theory of rate coding, the second one emphasizes the function of temporal precision in the neural code. Since then, numerous studies have demonstrated the emergence of synchronicity of activity within a neuronal population~\citep{riehle_spike_1997, davis_spontaneous_2021}, it's use for efficient coding thanks to the use of spike latencies~\citep{perrinet_coding_2004, gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2002, carr_circuit_1990}. All these findings and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and suggest the existence of spatio-temporal spiking motifs in biological spike trains. In neural models, the integration of heterogeneous delays~\citep{guise_bayesian_2014, zhang_supervised_2020, nadafian_bio-plausible_2020} allows for the efficient exploitation or detection of these spatio-temporal motifs embedded in the spike train. In particular,~\citep{izhikevich_polychronization_2006} introduced the notion of the polychronous group as a repetitive motif of spikes defined by a subset of neurons with different, but precise, relative spiking delays. Due to the variety of configurations and the possible coexistence of multiple superimposed motifs, this representation has a much higher information capacity than a firing rate-based approach to neural coding.

The present paper proposes additional experiments extending a recent model of spiking neurons with heterogeneous synaptic delays (i.e., the time between the arrival of an afferent spike at a given synapse and the contribution of the associated postsynaptic potential to the neuron's soma)~\cite{grimaldi_learning_2022}. This particular model was trained to solve a motion detection task on a synthetic event-based dataset generated by moving parameterized textures. Once trained, the volume of event-driven computations could be drastically reduced by pruning the synapses, while maintaining top performance for the classification task. This was a demonstration that formal neurons can exploit the precise timing of spikes to improve neural computation. 
In the present work, we extend motion detection to a more ecological task. Instead of the synthetic textures that have been used to generate motion-driven stimuli, we use natural images and random movements that mimic biological saccades. 
Using an ecological motion detection task, we study the emergence of spatio-temporal spiking motifs when a single layer of spiking neurons is trained on a supervised classification task. First, we define the ecological cognitive task the model has to solve with the different datasets it will be tested on. Then, we develop a theoretically defined Heterogeneous Delays in Spiking Neural Network (HD-SNN) model capable of learning heterogeneous synaptic delays. We will investigate the efficiency of the motion detection mechanism and in particular its resilience to synaptic weight pruning. In this way, we will be able to show how such a model can provide an efficient solution to the trade-off between energy and accuracy.

>>>>

The response of a biological neuron depends on the precise timing of afferent spikes. This temporal aspect of the neural code is essential to understand information processing in neurobiology and applies particularly well to the output of neuromorphic hardware such as event-based cameras. However, most artificial neural models do not take advantage of this important temporal dimension of the neural code. Inspired by this neuroscientific observation, we develop a model for the efficient detection of temporal spiking motifs based on a layer of spiking neurons with heterogeneous synaptic delays, which we apply to the computer vision task of motion detection. Indeed, the diversity of synaptic delays on the dendritic tree allows for the synchronization of specific arrangments of synaptic inputs as they reach the basal dendritic tree. We show that this can be formalized as a time-invariant logistic regression that can be trained on labeled data. We apply this model to solve the specific computer vision problem of motion detection. We demonstrate its application to synthetic naturalistic videos transformed into event streams similar to the output of event-based cameras. In particular, we quantify how the accuracy of the model can vary as a function of the overall computational load. This end-to-end, event-driven computational building block could improve the performance of future spiking neural network (SNN) algorithms and in particular their implementation in neuromorphic chips.

@@

The human brain has the remarkable property of being able to react at any time while consuming a reasonable amount of energy, about tens of watts. This system is the result of millions of years of natural selection, and a striking difference between biological and artificial neural networks is the representation they use. Convolutional neural networks (CNNs), for example, represent the flow of information from one layer to another as tensors, storing visual information densely across the visual topography, with different properties represented in different channels. These networks are known to mimic many properties of biological systems, such that each can be attributed a ``Brain Score''~\citep{schrimpf_brain-score_2020}, yet this score takes no account of inference speed or energy consumption. CNNs have achieved state-of-the-art performance for some computer vision tasks, such as image recognition. However, they do not take advantage of the dynamics inherent in the way we perceive our natural environment. In the vast majority of biological neural networks, on the other hand, information is represented as \emph{spikes}, prototypical all-or-nothing events whose only parameters are their timing and the address of the neuron that fired the spike~\citep{paugam-moisy_computing_2012}. The third generation of artificial neural networks, known as spiking neural networks (SNNs), incorporates this temporal dimension into the way they perform their computations. They are of interest for computational neuroscience because they provide a better model of the biological neuron. Some approaches have developed normative models of SNNs that aim to have applications in machine learning. One of these is the SpikeNet algorithm, which uses a purely temporal approach by encoding information using one spike per neuron~\citep{delorme_spikenet_1999}. Another type of SNN that uses accurate timing of spikes attempts to determine the structure of the network to minimize a cost function. This was implemented in the SpikeProp algorithm~\citep{bohte_error-backpropagation_2002} and has been recently extended in gradient-based methods. This surrogate gradient method is now widely used in attempts to transfer the performance of CNNs to SNNs~\citep{zenke_remarkable_2021}. However, the performance of SNNs still lags behind that of firing rate based networks. The question of the advantage of using spikes in machine learning and computer vision remains open.

In a recent review paper, we reported experimental evidence for the presence of precise spiking motifs embedded in recordings from biological neural tissues~\cite{grimaldi_precise_2023}. These spatio-temporal motifs within the neural activity may be useful representations to perform computations for various cognitive tasks. In 1982, Abeles asked whether the role of cortical neurons is the integration of synaptic inputs or rather the detection of coincidences in temporal spiking motifs~\citep{abeles_role_1982}. While the first possibility favors the theory of rate coding, the second one emphasizes the function of temporal precision in the neural code. Since then, numerous studies have demonstrated the emergence of synchronicity of activity within a neuronal population~\citep{riehle_spike_1997, davis_spontaneous_2021}, it's use for efficient coding thanks to the use of spike latencies~\citep{perrinet_coding_2004, gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2002, carr_circuit_1990}. All these findings and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and suggest the existence of spatio-temporal spiking motifs in biological spike trains. In neural models, the integration of heterogeneous delays~\citep{guise_bayesian_2014, zhang_supervised_2020, nadafian_bio-plausible_2020} allows for the efficient exploitation or detection of these spatio-temporal motifs embedded in the spike train. In particular,~\citet{izhikevich_polychronization_2006} introduced the notion of the polychronous group as a repetitive motif of spikes defined by a subset of neurons with different, but precise, relative spiking delays, i.e., the time between the arrival of an afferent spike at a given synapse and the contribution of the associated postsynaptic potential to the neuron's soma. Due to the variety of configurations and the possible coexistence of multiple superimposed motifs, this representation has a much higher information capacity than a firing rate-based approach to neural coding.

The present paper proposes additional experiments extending a recent model of spiking neurons with heterogeneous synaptic delays~\cite{grimaldi_learning_2022}. This model was trained to solve a motion detection task on a synthetic event-based dataset generated by moving parameterized textures. Once trained, the volume of event-driven computations could be drastically reduced by pruning the synapses, while maintaining top performance for the classification task. This was a demonstration that formal neurons can exploit the precise timing of spikes to improve neural computations. In the present work, we extend motion detection to a more ecological task. Instead of the synthetic textures that have been used to generate motion-driven stimuli, we use natural images and random movements that mimic biological saccades. Using an ecological motion detection task, we study the emergence of spatio-temporal spiking motifs when a single layer of spiking neurons is trained on a supervised classification task. First, we define the ecological cognitive task the model has to solve with the different datasets it will be tested on. Then, we develop a theoretically defined Heterogeneous Delays Spiking Neural Network (HD-SNN) model capable of learning these heterogeneous synaptic delays. We will investigate the efficiency of the motion detection mechanism and in particular its resilience to synaptic weight pruning. In this way, we will be able to show how such a model can provide an efficient solution to the trade-off between energy and accuracy.
